---
title: "Analysis of 20,000 chess games from Lichess.com"
format: html
editor: visual
---

This is an analysis of over 20,000 online chess games played on Lichess.com. Data was downloaded from Kaggle.com. I will try to analyze whether white has an advantage, the median duration of games (time and no. of moves) and most popular openings played.

```{r}
library(tidyverse)
games <- read_csv("games.csv")
```

```{r}
winners_colour <- games |> 
  count(winner) |> 
  mutate(winner = factor(winner, levels = c("white", "black", "draw")))

winners_colour |> 
  ggplot(aes(x = winner, y = n)) +
  geom_col()

```

White has a slight advantage.

```{r}
games <- games |> 
  mutate(duration_minutes = (last_move_at - created_at) / (1000 * 60))

zero_duration_count <- games |> 
  filter(duration_minutes == 0) |> 
  count() |> 
  pull(n)

total_games <- games |> 
  count() |> 
  pull(n)

percentage_zero_games <- (zero_duration_count/total_games) * 100

print(percentage_zero_games)
```

There were a high percentage of games showing a play duration of zero. This is clearly an error in data collection, and those games will be removed for the rest of the analysis.

```{r}

games_greater_than_0 <- games |> 
  filter(duration_minutes != 0)

games_greater_than_0 |> 
  filter(duration_minutes != 0 & duration_minutes < 200) |> 
  ggplot(aes(x = duration_minutes)) +
  geom_histogram(binwidth = 1)



```

Most games are less than 50 minutes, but there is also a high number of games lasting for 166 minutes. This is probably because of a default system setting that records disconnected games as lasting for that long.

```{r}
games_greater_than_0 |> 
  filter(duration_minutes < 50) |> 
  ggplot(aes(x = duration_minutes)) +
  geom_histogram(binwidth = 1)
```

Zooming in, most games are about 10 minutes long.

```{r}
games_greater_than_0 <- games_greater_than_0 |> 
  mutate(move_count = str_count(moves, "\\S+"))

games_greater_than_0 |> 
  filter(move_count < 250) |> 
  ggplot(aes(x = move_count)) +
  geom_histogram(binwidth = 1)

```

As for the moves, most games last for about 50 moves, but there is a strange spike at the beginning.

```{r}
n_games_with_x_moves <- games_greater_than_0 |> 
  count(move_count)

victory_status_counts <- games_greater_than_0 |> 
  filter(move_count == 2) |> 
  count(victory_status)
  
print(victory_status_counts)

```

153 games end in two moves, all by resignation or running out of time.

```{r}

grouped_by_opening <- games_greater_than_0 |> 
  group_by(opening_name) |> 
  summarise(game_count = n()) |> 
  arrange(desc(game_count))

print(grouped_by_opening, width = Inf)
```

These are the most popular openings played.

```{r}

popular_openings <- games_greater_than_0 |> 
  group_by(opening_name) |> 
  filter(n() >= 100) |> 
  ungroup()

black_win_rate <- popular_openings |> 
  filter(winner %in% c("black", "white")) |> 
  group_by(opening_name, winner) |> 
  summarise(game_count = n(), .groups = "drop") |> 
  pivot_wider(names_from = winner, values_from = game_count, values_fill = 0) |> 
  rename(black_win_count = black, white_win_count = white) |> 
  arrange(desc(black_win_count + white_win_count))

black_win_rate <- black_win_rate |> 
  mutate(black_win_rate = black_win_count / (black_win_count + white_win_count) * 100) |> 
  arrange(desc(black_win_rate))
  


```
